---
title: "Diabetes and Obesity:Trends & Insights"
author: "Kayla Lu"
date: "2026-04-03"
output: html_document
---

# Introduction
Diabetes is a critical global health priority, as its rising prevalence places an increasing burden on healthcare systems and individual quality of life. 

In this analysis, we examine the diabetes prevalence in the top 20 countries and track how the rates changed from 2011 to 2021. By comparing this data with national obesity rates, the project uses exploratory data analysis to highlight the relationship between weight-related metrics and the decade-long growth of the diabetes epidemic.

# Data Preparation
```{r setup, message=FALSE, warning=FALSE}
# Load required packages
library(tidyverse)
library(ggplot2)

# Read dataset
diabetes <- read.csv("~/Desktop/R project/3_Diabetes/Diabetes prevalence.csv")
str(diabetes)

obesity <- read.csv("~/Desktop/R project/3_Diabetes/Obesity prevalance.csv")
str(obesity)
```

## Data Inspection
```{r}
head(diabetes)
str(diabetes)
summary(diabetes)
colSums(is.na(diabetes))

head(obesity)
str(obesity)
summary(obesity)
colSums(is.na(obesity))
```

## Data Cleaning/ Reshaping
```{r}
diabetes_1 <- diabetes %>%
  select(REF_AREA, REF_AREA_LABEL, X2011, X2021) %>%
  mutate(diabetes_change = X2021 - X2011) %>%
  rename(Location = REF_AREA_LABEL,
         diabetes_2021 = X2021,
         diabetes_2011 = X2011,
         code = REF_AREA)

obesity_1 <- obesity %>%
  select(SpatialDimValueCode, Location, Period, Value) %>%
  pivot_wider(
    names_from = Period,
    values_from = Value
    ) %>%
  rename(obesity_2021 = "2021",
         obesity_2011 = "2011",
         code = SpatialDimValueCode) %>%
  mutate(
    obesity_2011 = as.numeric(str_extract(obesity_2011, "^[0-9.]+")),
    obesity_2021 = as.numeric(str_extract(obesity_2021, "^[0-9.]+")),
    obesity_change = obesity_2021 - obesity_2011
  )
```

# Exploratory Data Analysis
## Global Diabetes Prevalance Average 
```{r}
diabetes_summary_avg <- diabetes_1 %>%
  summarise(
    dm_mean_2011 = mean(diabetes_2011, na.rm = TRUE),
    dm_mean_2021 = mean(diabetes_2021, na.rm = TRUE),
    dm_mean_change = mean(diabetes_change, na.rm = TRUE)
  )
print(diabetes_summary_avg)
```

The global mean diabetes prevalence rose from 8.01% in 2011 to 9.07% in 2021.

## Diabetes Prevalance Distribution 
```{r}
diabetes_long <- diabetes_1 %>%
  select(Location, diabetes_2011, diabetes_2021) %>%
  pivot_longer(cols = starts_with("diabetes"), 
               names_to = "Year", 
               values_to = "Prevalence") %>%
  mutate(Year = ifelse(Year == "diabetes_2011", "2011", "2021")) %>%
  filter(!is.na(Prevalence))

ggplot(diabetes_long, aes(Year, Prevalence)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Diabetes Prevalence (2011 vs. 2021)", y = "Prevalence (%)", x = NULL) +
  theme_minimal()
```

The boxplot shows a clear upward shift in global diabetes prevalence from 2011 to 2021, with both the median and the overall distribution reaching higher levels. The increased number of outliers in 2021 highlights that more countries are facing extreme prevalence rates, some now exceeding 30%

## Diabetes Prevalence Top20 countries (2021)
```{r}
top20_DM_2021 <- diabetes_1 %>%
  arrange(desc(diabetes_2021)) %>%
  head(20)

ggplot(top20_DM_2021, aes(x = reorder(Location, diabetes_2021), y = diabetes_2021)) +
  geom_col(fill = "skyblue") +
  geom_text(aes(label = round(diabetes_2021,1)), 
            hjust = -0.1) +
  labs(title = "Top 20 Countries by Diabetes Prevalence in 2021",
       subtitle = "Prevalence among adults aged 20–79",
       x = NULL, y = "Prevalance(%)") +
  coord_flip() +
  theme_minimal()
```

Pakistan leads with a 30.8% prevalence, significantly outperforming other nations on the list. The top 20 is heavily concentrated in Pacific Island nations and the Middle East.

## Diabetes Prevalence Change Overview
```{r}
diabetes_1 %>%
  filter(!is.na(diabetes_change)) %>%
  ggplot(aes(y = diabetes_change)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Distribution of Prevalence Change (2011-2021)", 
       y = "Change in Prevalence (%)", x = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

The prevalence change boxplot shows that most countries experienced a modest increase, while several extreme outliers saw spikes of over 20 percentage points.

## Diabetes Prevalence Change Top20 countries
```{r}
top20_DMchange <- diabetes_1 %>%
  arrange(desc(abs(diabetes_change))) %>%
  head(20)

ggplot(top20_DMchange, aes(x = reorder(Location, abs(diabetes_change)), y = diabetes_change)) +
  geom_col(fill = "skyblue") +
  geom_text(aes(label = diabetes_change), hjust = ifelse(top20_DMchange$diabetes_change > 0, -0.1, 1.1)) +
  coord_flip() +
  labs(
    title = "Top 20 Countries by Diabetes Prevalence Change (2011–2021)",
    subtitle = "Prevalence among adults aged 20–79",
    y = "Prevalence Change (%)",
    x = NULL
  ) +
  theme_minimal()
```

While Pakistan shows a staggering 22.9% increase, the chart also reveals significant decreases in countries like Lebanon and Bahrain.

## Global Obesity Prevalence Average
```{r}
obesity_summary_avg <- obesity_1 %>%
  summarise(
    obesity_mean_2011 = mean(obesity_2011, na.rm = TRUE),
    obesity_2021 = mean(obesity_2021, na.rm = TRUE),
    obesity_change = mean(obesity_change, na.rm = TRUE)
  )
print(obesity_summary_avg)
```
The obesity rate saw a significant 4.8 percentage point increase over ten years, growing from 19.4% to 24.2%.

## Obesity Prevalance Distribution 
```{r}
obesity_long <- obesity_1 %>%
pivot_longer(
    cols = c(obesity_2011, obesity_2021),
    names_to = "Year",
    values_to = "Prevalence"
  ) %>%
  mutate(Year = str_replace(Year, "obesity_", "")) %>%
  filter(!is.na(Prevalence))

ggplot(obesity_long, aes(Year, Prevalence)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Obesity Prevalence (2011 vs. 2021)", y = "Prevalence (%)", x = NULL) +
  theme_minimal()
```

The boxplot shows a clear upward shift in global obesity rates from 2011 to 2021. Both the median and overall distribution have moved higher.

## Obesity Prevalence Change Overview
```{r}
obesity_1 %>%
  filter(!is.na(obesity_change)) %>%
  ggplot(aes(y = obesity_change)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Distribution of Prevalence Change (2011-2021)", 
       y = "Change in Prevalence (%)", x = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

The median increase in obesity is around 4.7%, showing a universal upward trend. Most regions experienced a growth of 3% to 7% over the decade.

## Obesity Prevalence Change Top20 countries (2021)
```{r}
top20_Obesitychange <- obesity_1 %>%
  arrange(desc(abs(obesity_change))) %>%
  head(20)

ggplot(top20_Obesitychange, aes(x = reorder(Location, abs(obesity_change)), y = obesity_change)) +
  geom_col(fill = "skyblue") +
  geom_text(aes(label = obesity_change), hjust = ifelse(top20_Obesitychange$obesity_change > 0, -0.1, 1.1)) +
  coord_flip() +
  labs (
    title = "Top 20 Countries by Obesity Prevalance Change (2011–2021)",
    subtitle = "Prevalence among adults (18 years and older)",
    y = "Prevalence Change (%)",
    x = NULL
  ) +
  theme_minimal()
```

Romania recorded the largest increase in obesity prevalence, rising by over 13%. Several other nations, including Uzbekistan and Pakistan, also saw significant growth exceeding 9%.

## Compare Diabetes and Obesity Prevalence
### Merge dataset
```{r}
data <-obesity_1 %>%
  left_join(diabetes_1, by = "code")

data <- data %>%
  select(
    code,
    Country = Location.x,
    obesity_2011, obesity_2021, obesity_change,
    diabetes_2011, diabetes_2021, diabetes_change
  )

nrow(obesity_1)
nrow(data)

colSums(is.na(data))
data %>% 
  summarise(
    avg_obesity_2021 = mean(obesity_2021, na.rm = TRUE),
    avg_diabetes_2021 = mean(diabetes_2021, na.rm = TRUE)
  )

missing_list <- data %>%
  filter(is.na(diabetes_change)) %>%
  select(code, Country)
print(missing_list)

data_clean <- data %>%
  drop_na(obesity_change, diabetes_change)
```

## Correlation and Regression Analysis: Obesity vs. Diabetes (2021)
```{r}
cor.test(data_clean$diabetes_2021, data_clean$obesity_2021)

model_2021 <- lm(diabetes_2021 ~ obesity_2021, data = data_clean)
summary(model_2021)

ggplot(data_clean, aes(x = obesity_2021, y = diabetes_2021)) +
  geom_point() +
  labs(title = "Relationship between Obesity and Diabetes Prevalence (2021)",
       x = "Obesity prevalence",
       y = "Diabetes prevalence"
         ) +
  geom_smooth(method = "lm")
```

The upward-sloping regression line shows a positive correlation between obesity and diabetes in 2021. However, significant outliers indicate that other factors also influence diabetes rates.

## Correlation between Obesity and Diabetes Trends (2011–2021)
```{r}
cor.test(data_clean$obesity_change, data_clean$diabetes_change)

model_change <- lm(diabetes_change ~ obesity_change, data = data_clean)
summary(model_change)

ggplot(data_clean, aes(x = obesity_change, y = diabetes_change)) +
  geom_point() +
  labs(title = "Relationship between Obesity and Diabetes Prevalence Change (2011-2021)",
       x = "Obesity prevalence change",
       y = "Diabetes prevalence change"
         ) +
  geom_smooth(method = "lm")
```

The flat regression line indicates a weak correlation between 10-year changes in obesity and diabetes. This suggests that national diabetes trends are driven by factors beyond just obesity growth.

## Finding Exceptions
```{r}
data_clean <- data_clean %>%
  mutate(res_2021 = residuals(model_2021))
```

## Identifying Positive Outliers: Higher Than Predicted
```{r}
top_positive_res <- data_clean %>%
  arrange(desc(res_2021)) %>%
  head(5) %>%
  select(Country, obesity_2021, diabetes_2021, res_2021)
print(top_positive_res)
```
These nations show significantly higher diabetes rates than predicted by their obesity levels. Pakistan shows the highest positive residual, with diabetes rates soaring far beyond what its obesity levels would suggest.

## Identifying Negative Outliers: Lower Than Predicted
```{r}
top_negative_res <- data_clean %>%
  arrange(res_2021) %>%
  head(5) %>%
  select(Country, obesity_2021, diabetes_2021, res_2021)
print(top_negative_res)
```

These nations show significantly lower diabetes rates than predicted by their obesity levels. Samoa and Ireland exhibit the largest negative residuals.

# Conclusion
Our analysis confirms that the **rising prevalence of diabetes and obesity** is indeed a significant global burden. However, the intensity of this burden varies considerably across the countries studied.

By tracking data from 2011 to 2021, we found that while national obesity rates are strong indicators of current diabetes prevalence, they **do not fully explain the decade-long growth** observed in every nation. This suggests that the diabetes epidemic is driven by a complex interplay of factors, including genetics, dietary habits, and socioeconomic conditions.
